---
title: "Covariance adjustment that stays exact"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariance adjustment that stays exact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(20260624)
```

## The temptation, and the trap

You measured covariates that predict the outcome. It is tempting to use them: fit
a model, subtract its prediction, and test the residuals, which vary less than the
raw outcome, so the test should have more power. The temptation is real and the
gain can be large. The trap is that the obvious way of doing it can quietly
invalidate the test.

The obvious way is to fit the model once on the observed data, form the residuals,
and then permute treatment among those fixed residuals. When the model overfits,
the fitted values chase the observed treatment assignment, the residuals carry a
trace of it, and the permutation reference no longer matches the test statistic.
The test rejects too often under the null --- you buy power by spending your error
rate without meaning to.

riposte avoids this. It adjusts for covariates in a way that keeps the test
*exactly* valid under the sharp null, for any predictive model you care to use ---
ridge regression, a random forest, anything measurable --- and at any number of
re-randomizations.

## How riposte adjusts

Two rules make the difference:

1. **Fit on the controls only.** The model never sees the treated outcomes, so it
   cannot absorb the treatment effect into its predictions.
2. **Refit inside every re-randomization.** Each time treatment is re-randomized,
   the model is fit afresh on that re-randomization's control units, and the
   residuals are recomputed.

The second rule is the one that matters for validity. Under the sharp null the
outcomes are fixed, so refitting makes the entire procedure --- model fit,
residualization, combination --- a deterministic function of the assignment. A
deterministic function of the assignment has an exact randomization distribution.
This is the design-based guarantee in the spirit of Rosenbaum: the test statistic
is the whole algorithm, the refit included, so its permutation distribution is
exact regardless of how the model behaves. (The argument and its scope are in the
package's development note; the method is from Bowers, Fredrickson, Hansen, and
Yin, "Regression Without Regrets.")

You pass the covariates with `adjust`; the learner defaults to ridge.

```{r basic}
library(riposte)

# a block-randomized experiment with predictive covariates and a small effect
B <- 10L; nb <- 12L
block <- factor(rep(seq_len(B), each = nb)); N <- B * nb
z <- integer(N)
for (b in levels(block)) { ix <- which(block == b); z[ix][sample.int(nb, nb %/% 2L)] <- 1L }
X <- matrix(rnorm(N * 3), N, 3)
y <- as.numeric(X %*% c(2, -1.5, 1)) + rnorm(N, sd = 0.8) + 0.5 * z  # covariates predict y
dat <- data.frame(Y = y, trt = z, blk = block, x1 = X[, 1], x2 = X[, 2], x3 = X[, 3])

unadjusted <- riposte_test(Y ~ trt | blk, dat, statistic = "max",
                           representations = list(raw = function(y) y),
                           nresample = 999, seed = 1)
adjusted <- riposte_test(Y ~ trt | blk, dat, statistic = "max",
                         representations = list(raw = function(y) y),
                         adjust = ~ x1 + x2 + x3, nresample = 999, seed = 1)
c(unadjusted = unadjusted$p.value, adjusted = adjusted$p.value)
```

When the covariates predict the outcome, the adjusted p-value is the smaller one:
removing the covariates' variance sharpens the comparison. When they do not
predict, adjustment costs almost nothing, and either way the p-value stays valid.

## The fit-once trap, made visible

To see why the refit matters, compare it to the fit-once shortcut
(`adjust_refit = FALSE`) under the sharp null, with a deliberately overfit learner
(ordinary least squares with 20 covariates fit on 30 controls). We simulate the
rejection rate when there is *no effect*: a valid test rejects about 5% of the
time. The simulation is small (40 replications) so the vignette builds quickly;
the contrast is stark enough to be unmistakable anyway.

```{r trap}
size <- function(refit, nsims = 40, nres = 49, d = 20) {
  Bs <- 6L; nbs <- 10L; Ns <- Bs * nbs
  blk <- factor(rep(seq_len(Bs), each = nbs))
  fml <- stats::as.formula(paste0("~", paste0("x", seq_len(d), collapse = "+")))
  p <- numeric(nsims)
  for (i in seq_len(nsims)) {
    Xs <- matrix(rnorm(Ns * d), Ns, d)
    ys <- rnorm(Ns)                                   # sharp null: no effect, X unrelated
    zs <- integer(Ns)
    for (b in levels(blk)) { ix <- which(blk == b); zs[ix][sample.int(nbs, nbs %/% 2L)] <- 1L }
    d0 <- data.frame(Y = ys, trt = zs, blk = blk, Xs); names(d0)[-(1:3)] <- paste0("x", seq_len(d))
    p[i] <- riposte_test(Y ~ trt | blk, d0, statistic = "quadratic",
                         adjust = fml, learner = riposte_lm_learner(),
                         adjust_refit = refit, nresample = nres)$p.value
  }
  mean(p <= 0.05)
}

set.seed(1)
c(refit = size(TRUE), fit_once = size(FALSE))
```

The refit holds the level; the fit-once shortcut, with an overfit model, rejects
far more than 5% of the time (here most of the time --- the overfit residuals are
nearly collinear with the observed assignment). riposte uses the refit by default.
The shortcut is available, and honestly labelled, only for this comparison.

## What "exact" means here, precisely

The level is exact in finite samples: under the sharp null the test rejects with
probability at most the nominal level, for any model and any `nresample`. Two
things are worth stating plainly so the guarantee is not over-read.

- The p-value is a *Monte-Carlo* permutation p-value --- the share of the
  re-randomizations at least as extreme as observed, with the observed assignment
  included. That observed-inclusion convention is what makes the level exact for
  any number of draws; the p-value estimates, rather than equals, the
  full-enumeration value, so it varies a little with the seed.
- For the omnibus quadratic and the screen, riposte must estimate the covariance
  of the representations from the re-randomizations (the closed-form moments it
  uses without adjustment do not apply once the outcome is residualized afresh
  each draw). It estimates that covariance from the *pooled* set of statistics
  (observed and draws together) so that the metric is symmetric across
  assignments; this is what keeps the omnibus exact rather than anti-conservative.
  The estimation adds Monte-Carlo noise to the *metric*, which affects power, not
  level, and shrinks as `nresample` grows.

## Choosing a learner

The default is ridge regression, which has a closed form and shrinks its
coefficients, so it stays stable when there are many covariates --- the regime
where an unpenalized fit overfits. Any function of the form
`(X, Y, control) -> fitted values` works, so you can supply a random forest, a
LASSO, or your own:

```{r learner, eval = FALSE}
rf_learner <- function(X, Y, control) {
  fit <- randomForest::randomForest(X[control, , drop = FALSE], Y[control])
  as.numeric(predict(fit, X))
}
riposte_test(Y ~ trt | blk, dat, adjust = ~ x1 + x2 + x3, learner = rf_learner)
```

The exactness guarantee does not depend on the learner being any good. A poor
model simply yields a less powerful test, never an invalid one --- the regret-free
property the method is named for.

## Cluster-randomized designs

Adjustment composes with cluster randomization: pass `clusters`, and riposte
collapses to the cluster level (covariates averaged within cluster), fits the
learner on the control clusters, and refits over cluster re-randomizations.

```{r clusters, eval = FALSE}
riposte_test(Y ~ trt | blk, dat, clusters = "cluster_id",
             adjust = ~ x1 + x2 + x3, nresample = 999)
```

## When to reach for it

Use covariance adjustment when you have covariates that plausibly predict the
outcome and you want the extra power without giving up the validity of a
randomization test. The cost is computational --- the learner is refit once per
re-randomization --- and the only tuning that affects power (not validity) is the
choice of learner and, for the omnibus, the screen's shrinkage. Validity you get
for free.

## References

- Bowers, J., Fredrickson, M., Hansen, B., and Yin, X. Regression Without
  Regrets. In preparation.
- Rosenbaum, P. R. (2002). *Observational Studies*, 2nd ed. Springer. (Exact
  permutation inference under the sharp null.)
- Phipson, B. and Smyth, G. K. (2010). Permutation p-values should never be zero.
  *Statistical Applications in Genetics and Molecular Biology* 9(1).
