---
title: "Why the large-sample Cauchy combination is truncated at 0.9"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why the large-sample Cauchy combination is truncated at 0.9}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(riposte)
f <- function(x, d = 2) formatC(x, format = "f", digits = d)
sims <- readRDS(system.file("extdata", "truncation_sims.rds", package = "riposte"))
```

## A combined p-value of 1 next to a difference in means of 0.028

Wallsten and Nteta (2016) ran a survey experiment in which 36 respondents, 18
of them treated, answered on a 5-point scale coded 0, 0.25, 0.5, 0.75, and 1.
Sarkar and Coppock (2026) reanalyzed it with other experiments on religious
messages, and the cleaned data in their replication archive give these
counts:

```{r data}
vals <- c(0, 0.25, 0.5, 0.75, 1)
wal <- data.frame(Y = c(rep(vals, c(10, 2, 3, 2, 1)), rep(vals, c(5, 2, 0, 8, 3))),
                  Z = rep(0:1, each = 18))
table(answer = wal$Y, treated = wal$Z)
```

The difference in means gives a p-value of 0.028. riposte's Cauchy combination,
with its p-value from the large-sample formula that riposte used before
version 0.0.0.9008, gives exactly 1:

```{r puzzle}
t.test(Y ~ Z, data = wal)$p.value
riposte_test(Y ~ Z, data = wal, statistic = "cauchy", engine = "asymptotic",
             cauchy_truncation = 1)$p.value
```

A combined p-value of 1 says that the observed result is the least remarkable
result any assignment of treatment could have produced. This vignette shows
where that 1 comes from, how the truncated conversion of Gui, Jiang, and Wang
(2025) removes it, and why riposte uses their truncation level of 0.9 by
default. We will show that one of the six representations has a correct
p-value of exactly 1, that the untruncated Cauchy combination turns any p-value
of 1 into a combined p-value of 1, and that a smaller truncation level than 0.9
would trade a little more power for a few more false positives.

## The untruncated Cauchy combination lets one p-value of 1 decide

The Cauchy combination of Liu and Xie (2020) converts each representation's
p-value $p$ to $\tan((0.5 - p)\pi)$, averages the
converted values, and turns the average back into a p-value with the standard
Cauchy distribution, for which a variable $C$ exceeds $x$ with probability
$P(C > x) = 1/2 - \arctan(x)/\pi$. The
conversion makes a small p-value into a large number, so that one
representation with strong evidence is not diluted by others with none:

```{r conversion}
p <- c(0.001, 0.01, 0.05, 0.5, 0.95, 0.99, 0.999, 1)
data.frame(p = p, converted = format(tan((0.5 - p) * pi), digits = 4))
```

The conversion treats p-values near 1 the same way with the sign reversed:
0.999 becomes about $-318$, and 1 becomes minus infinity (R prints a huge
finite number because its value of $\pi/2$ is rounded). One converted value of
minus infinity makes the average minus infinity, and the combined p-value is
then 1 whatever the other p-values are. riposte's default representations are
six: the raw outcome, its rank, mean distance, mean rank distance, max
distance, and Huber (the outcome minus its median, divided by the median
of the absolute differences from the median, and capped at plus or minus
1.345). These are their six large-sample p-values in
the Wallsten and Nteta data:

```{r components}
riposte_test(Y ~ Z, data = wal, statistic = "cauchy", engine = "asymptotic")$component_p
```

The max distance's p-value is exactly 1. The max distance scores each answer by
its distance to the farther end of the observed range, which measures how far
the answer is from the middle of the scale. To compare the treated
respondents' average score with everyone's, we need each answer's score:

| answer       | 0 | 0.25 | 0.5 | 0.75 | 1 |
|--------------|---|------|-----|------|---|
| max distance | 1 | 0.75 | 0.5 | 0.75 | 1 |

 The 18 treated respondents
include 8 at an end and 10 one step in, so their average score is
$(8 \times 1 + 10 \times 0.75)/18 = 31/36$. All 36 include 19 at an end, 14 one
step in, and 3 in the middle, so their average is
$(19 \times 1 + 14 \times 0.75 + 3 \times 0.5)/36 = 31/36$. The treated average
equals the overall average, so the standardized difference is exactly 0, and
its two-sided p-value, the chance of a difference at least as far from 0, is 1.
That p-value is correct: on the max distance the arms do not differ at all. On
outcomes with few values and few units, a treated average equals the overall
average exactly in a noticeable share of assignments, and this can happen to
any representation.

## The truncated conversion of Gui, Jiang, and Wang

Gui, Jiang, and Wang (2025, Section 2.2) convert each p-value with the quantile
function of a Cauchy distribution cut off below. They keep a share $t$ of the
Cauchy distribution, above the point $c = \tan((0.5 - t)\pi)$. Write $F_C$
for the Cauchy distribution function, so $F_C(c) = 1 - t$. The cut-off
distribution has distribution function $F(x) = (F_C(x) - (1 - t))/t$ for
$x \ge c$. Its quantile for $p$ is the $x$ with $F(x) = 1 - p$, that is
$F_C(x) = 1 - tp$, and since the Cauchy quantile of $1 - q$ is
$\tan((0.5 - q)\pi)$, the conversion is $x = \tan((0.5 - tp)\pi)$. With $t = 0.9$, a p-value of 1 converts to
$\tan(-0.4\pi) = -3.08$ instead of minus infinity. The cut-off distribution's
upper tail is the Cauchy's divided by $t$. With $n$ p-values whose converted
values sum to $S$, Gui, Jiang, and Wang take the combined p-value to be
$\min(1, n P(C > S)/t)$, with $C$ a standard Cauchy variable (their Section
2.3). The reasoning is that for variables with heavy upper tails, a large sum
almost always comes from one large term, so the chance that the sum of $n$
converted values exceeds $S$ is about $n$ times the chance that one of them
does, and one converted value exceeds $S$ with probability $P(C > S)/t$. With one p-value
this returns that p-value exactly. `riposte_truncated_cauchy()` computes it,
and `riposte_test()` uses it for the large-sample Cauchy combination and hybrid:

```{r truncated}
riposte_test(Y ~ Z, data = wal, statistic = "cauchy", engine = "asymptotic")$p.value
```

The max distance's p-value of 1 now converts to $-3.08$ instead of minus
infinity, and the raw outcome, rank, and Huber p-values near 0.03 still
count. Fang, Chang, Park, and
Tseng (2023) proposed an earlier truncation of the same kind, replacing every
p-value above 0.99 with 0.99, and noted that p-values near 1 are common with
discrete data.

## What the truncation level changes

The choice of $t$ is easiest to see in the rule for rejecting. With six
p-values, the combined p-value is at most 0.05 exactly when
$P(C > S) \le 0.05t/6$, that is, when $S \ge \cot(\pi \cdot 0.05t/6)$, since
$P(C > x) = q$ exactly when $x = \cot(\pi q)$. Since each converted value is
$\cot(\pi t p_k)$, multiplying both sides by $\pi t$ gives
$\sum_k \pi t \cot(\pi t p_k) \ge \pi t \cot(\pi t \cdot 0.05/6)$. We call
$\pi t \cot(\pi t p)$ the contribution of a p-value $p$. The rule is then:
reject when the six contributions add up to at least the contribution of a
single p-value of $0.05/6$. Because $\pi t \cot(\pi t p)$ is close to $1/p$
for small $p$, that threshold is close to $6/0.05 = 120$ for any $t$.

```{r contributions}
contribution <- function(p, t) pi * t / tan(pi * t * p)
p <- c(0.01, 0.05, 0.5, 0.8, 1)
round(cbind(p, `t = 0.9` = contribution(p, 0.9), `t = 0.8` = contribution(p, 0.8),
            `t = 0.5` = contribution(p, 0.5), `1/p` = 1 / p), 2)
```

A small p-value contributes about $1/p$ whatever $t$ is, so $t$ changes only
what large p-values contribute. At $t = 0.9$ a p-value of 1 subtracts 8.7, at
$t = 0.8$ it subtracts 3.5, and at $t = 0.5$ it subtracts nothing. As $t$
approaches 0 every p-value contributes $1/p$, and the rule becomes
$\sum_k 1/p_k \ge 6/0.05$, that is, $6/\sum_k (1/p_k) \le 0.05$. That is the
harmonic mean p-value of Wilson (2019): the number of p-values divided by the
sum of their reciprocals. Lowering $t$ only raises the contributions of
large p-values, so a smaller $t$ rejects in every experiment that a larger $t$
rejects, and in some more. Power and the false positive rate therefore rise
together as $t$ falls, and the smallest p-value over several values of $t$ is
always the p-value at the smallest of them.

## How much power, and how many false positives

`dev/truncation-sims.R` in the package source measures both. For false
positives it holds outcomes fixed, so that the treatment changes nothing, and
reassigns treatment `r format(sims$draws[1], big.mark = ",")` times: once for
the Wallsten and Nteta data and once for 40 outcomes drawn from a skewed
(lognormal) distribution, half treated. For power it draws
`r format(sims$draws[3], big.mark = ",")` experiments with 40 lognormal
outcomes, half treated, in which the treatment either adds 0.55 to every
treated outcome or moves treated outcomes 2.1 times as far from the control
median. Every rule is applied to the same six large-sample p-values. Each entry
is the share of experiments rejected at 0.05:

```{r rates, echo = FALSE}
knitr::kable(sims$rates, digits = 3)
```

The standard error of a rate near 0.05 from
`r format(sims$draws[1], big.mark = ",")` reassignments is
$\sqrt{0.05 \times 0.95 / `r sims$draws[1]`} = `r f(sqrt(0.05 * 0.95 / sims$draws[1]), 4)`$,
and of a power near 0.5 from `r format(sims$draws[3], big.mark = ",")`
experiments about `r f(sqrt(0.25 / sims$draws[3]), 3)`. The six p-values in
these designs are correlated. With six independent p-values, the case in
which these rules exceed 0.05 by the most in Gui, Jiang, and Wang's
Figure 2, one million simulated sets give these false positive rates:

```{r independent, echo = FALSE}
knitr::kable(data.frame(rule = names(sims$independent), `false positive rate` = sims$independent,
                        check.names = FALSE, row.names = NULL), digits = 4)
```

The untruncated combination rejects too rarely in the Wallsten data, because a
p-value of exactly 1 occurs in many reassignments and forces the combined
p-value to 1 each time. Truncating at 0.9 brings the rate close to 0.05 and
raises power. Moving from 0.9 to smaller levels adds about a point of power at
most, while the false positive rate keeps rising. With independent p-values it
is already `r f(sims$independent[["t = 0.9"]], 3)` at 0.9 and
`r f(sims$independent[["t = 0.5"]], 3)` at 0.5. Gui, Jiang, and Wang recommend 0.9 from their
own simulations (their Section 6), and riposte uses it.

## Several truncation levels at once

Since the smallest p-value over several truncation levels is the p-value at the
smallest level, choosing the best level after seeing the data is the same as
always using the smallest, which has the most false positives above. Taking
the p-value from re-randomization removes that problem. When the treatment
changed no one's outcome, the observed assignment is one more random draw from
the assignment process, so its statistic is no more likely than any
reassignment's to be the most extreme. A p-value that counts the share of
reassignments at least as extreme as the observed statistic is then at most
0.05 no more than 5% of the time, whatever the statistic is. So the simulation
also took each statistic's p-value from `r sims$B` reassignments, in
`r format(sims$several_draws, big.mark = ",")` experiments per column:

```{r several, echo = FALSE}
several <- sims$several
rownames(several) <- c("t = 0.9, large-sample", "smallest over t = 0.5, 0.7, 0.9, large-sample",
                       "t = 0.9, re-randomization", "smallest over t = 0.5, 0.7, 0.9, re-randomization")
knitr::kable(several, digits = 3)
```

The standard errors here are about 0.005 for the false positive rate and 0.011
for power. Under re-randomization, one level and several levels reject at
nearly the same rates, so using several levels gains nothing. Taking the
p-value from re-randomization rather than from the large-sample formula raised
power by about 3 percentage points at $t = 0.9$.

## The re-randomization Cauchy combination

With the default `engine = "permute"`, riposte gives each representation a
mid-p value: the share of reassignments whose statistic is more extreme than
the observed one, counting reassignments tied with it as one half. A mid-p
value is never exactly 1. riposte averages the converted mid-p values and
takes the share of reassignments whose average is at least the observed one.
For the reason given in the previous section, the test's false positive rate
is then at most the nominal level whatever the conversion. `cauchy_truncation` therefore applies only to
`engine = "asymptotic"`.

## References

Fang, Y., Chang, C., Park, Y., and Tseng, G. C. (2023). Heavy-tailed
distribution for combining dependent p-values with asymptotic robustness.
*Statistica Sinica*, 33, 1115-1142. doi:10.5705/ss.202022.0046

Gui, L., Jiang, Y., and Wang, J. (2025). Aggregating dependent signals with
heavy-tailed combination tests. *Biometrika*, 112(4), asaf038.
doi:10.1093/biomet/asaf038

Liu, Y., and Xie, J. (2020). Cauchy combination test: a powerful test with
analytic p-value calculation under arbitrary dependency structures. *Journal
of the American Statistical Association*, 115(529), 393-402.
doi:10.1080/01621459.2018.1554485

Long, M., Li, Z., Zhang, W., and Li, Q. (2023). The Cauchy combination test
under arbitrary dependence structures. *The American Statistician*, 77(2),
134-142. doi:10.1080/00031305.2022.2116109

Sarkar, R., and Coppock, A. (2026). The effects of religious messages and
endorsements on political attitudes: a meta-reanalysis. *American Political
Science Review*, advance online publication. doi:10.1017/S0003055426101695.
Replication archive: doi:10.7910/DVN/NHDFFS

Wallsten, K., and Nteta, T. M. (2016). For you were strangers in the land of
Egypt: clergy, religiosity, and public opinion toward immigration reform in the
United States. *Politics and Religion*, 9(3), 566-604.
doi:10.1017/S1755048316000444

Wilson, D. J. (2019). The harmonic mean p-value for combining dependent tests.
*Proceedings of the National Academy of Sciences*, 116(4), 1195-1200.
doi:10.1073/pnas.1814092116
